Amphistomy increases leaf photosynthesis more in coastal than montane plants of Hawaiian ʻilima (\emph{Sida fallax})

### 2023-11-03 ms/ms.tex
revised ms

\begin{document}
\maketitle
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, frame hidden, interior hidden, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, breakable, enhanced]}{\end{tcolorbox}}\fi

@@

\textbf{Key results} Coastal ʻilima leaves benefit 4 times more from
amphistomy compared to their montane counterparts. Our evidence was
equivocal with respect to two hypotheses -- that coastal leaves benefit
more because 1) they are thicker and therefore have lower CO\(_2\)
conductance through the internal airspace, and 2) that they benefit more
because they have similar conductance on each surface, as opposed to
most of the conductance being on the lower (abaxial) surface.

\textbf{Conclusions} This is the first direct experimental evidence that
amphistomy \emph{per se} increases photosynthesis, consistent with the
hypothesis that parallel pathways through upper and lower mesophyll
increase the supply of CO\(_2\) to chloroplasts. The prevalence of
amphistomatous leaves in open, sunny habitats can partially be explained
the increased benefit of amphistomy in ʻsunʻ leaves, but the mechanistic
basis of this observation is an area for future research.

@@

\autoref{fig:ags-curve} illustrates our method for collecting
\(A \textendash g_\text{sw}\) curves. We collected two curves per leaf,
an amphi (untreated) curve and a pseudohypo (treated) curve. To control
for order effects, we alternated between starting with amphi or
pseudohypo leaf measurements. In the field, we acclimated the focal leaf
to saturating light and high relative humidity (\(\mathrm{RH} = 70\%\)),
as described above, until \(A\) and \(g_\text{sw}\) reach their maximum.
We used these data as our estimates of maximum \(A\) and
\(g_\text{sw}\). After that, we decreased \(\mathrm{RH}\) to
\(\approx 10\%\) to induce rapid stomatal closure without biochemical
downregulation. Hence, \(A_\text{amphi}\) and \(A_\text{hypo}\) were
both measured at low chamber humidity after the leaf had acclimated to
high humidity. All other environmental conditions in the leaf chamber
remained the same. We logged data until \(g_\text{sw}\) reached its
nadir. We then repeated the process of acclimating the leaf to 70\%
\(\mathrm{RH}\) and inducing stomatal closure with low \(\mathrm{RH}\)
with the other treatment (amphi or pseudohypo). We did not detect an
effect of treatment order on \(\mathrm{AA}\) (results not shown).

@@

In 10 of 12 leaves, the minimum \(g_\text{sw}\) of the amphi curve was
smaller than the maximum \(g_\text{sw}\) of the pseudohypo curve
(i.e.~the curves overlapped for a range of \(g_\text{sw}\) values). In
those cases, we estimated \(\hat{A}_\text{amphi}\) and
\(\hat{A}_\text{hypo}\) at the \(g_\mathrm{sw}\) value in the middle of
the range of overlap between the curves. In 2 of 12 leaves, the
\(A \textendash g_\text{sw}\) curves did not quite overlap because the
minimum \(g_\text{sw}\) of the amphi curve was slightly greater than the
maximum \(g_\text{sw}\) of the pseudohypo curve. In those cases, we
estimated \(\mathrm{AA}\) by extrapolating slightly,
\(1.98\times 10^{-3}~\text{and}~3.29\times 10^{-3}~\text{mol}~\text{m}^{-2}~\text{s}^{-1}\),
beyond the measured curves to the \(g_\mathrm{sw}\) value in between the
curves. The vertical lines in \autoref{fig:licor} show the
\(g_\text{sw}\) for each leaf. \(\mathrm{AA}\) for each leaf was
calculated \(\hat{A}_\text{amphi}\) and \(\hat{A}_\text{hypo}\) using
the log-response ratio shown above.

To estimate \(\hat{\beta}\)'s from the \(A \textendash g_\text{sw}\)
curve for each leaf, we fit Bayesian regressions using the \emph{R}
package \textbf{brms} version 2.20.1 (Bürkner, 2017) with MCMC sampling
in \emph{Stan} (Stan Development Team, 2023). We used CmdStan version
2.33.0 and \textbf{cmdstanr} version 0.6.1 (Gabry and Češnovar, 2023) to
interface with \emph{R} version 4.3.1 (R Core Team, 2023). We sampled
the posterior distribution from 4 chains with 1000 iterations each after
1000 warmup iterations per chain. We estimated parameters and confidence
intervals as the median and 95\% quantile intervals of the posterior,
respectively. The key prediction is that
\(\mathrm{AA}_\text{coastal} > \mathrm{AA}_\text{montane}\), meaning the
95\% confidence intervals of
\(\mathrm{AA}_\text{coastal} - \mathrm{AA}_\text{montane}\) should be
positive and not encompass 0.

@@

We tested whether leaf thickness (log-transformed) varied between
coastal and montane populations and among individuals within populations
using a Bayesian mixed-effects model with habitat as a fixed effect and
individual plant and site as random effects. We used the \emph{R}
package \textbf{brms} version 2.20.1 (Bürkner, 2017) to fit the model in
\emph{Stan} (Stan Development Team, 2023) with CmdStan version 2.33.0
and \textbf{cmdstanr} version 0.6.1 (Gabry and Češnovar, 2023). We
sampled the posterior distribution from 4 chains with 1000 iterations
each after 1000 warmup iterations per chain. We estimated the
relationship between population average leaf thickness and
\(\mathrm{AA}\) measured from a single individual per population. We
used this approach because most of the variation in leaf thickness was
among sites and the plant selected for gas exchange measurements was not
always among the plants randomly selected for leaf thickness, precluding
individual level correlation. We propagated uncertainty about in
\(\mathrm{AA}\) and leaf thickness estimates by integrating over the
entire posterior distribution sample for each variable. The key
prediction is that the effect of leaf thickness on \(\mathrm{AA}\) is
positive, meaning the 95\% confidence interval of the slope should be
positive and not encompass 0.

@@

We tested whether \(g_\mathrm{smax,ratio}\) varied between coastal and
montane populations and among individuals within populations using a
Bayesian mutliresponse, mixed-effects model. The modeled response
variables are stomatal count and guard cell length on each surface.
Counts were modeled as negative binomially distributed variable from a
latent stomatal density and a parameter \(\phi\) to estimate
overdispersion in counts relative to a Poisson model. For all traits,
the explanatory variables were habitat as a fixed effect and leaf within
individual plant, individual plant, and site as random effects. We used
the \emph{R} package \textbf{brms} version 2.20.1 (Bürkner, 2017) to fit
the model in \emph{Stan} (Stan Development Team, 2023) with CmdStan
version 2.33.0 and \textbf{cmdstanr} version 0.6.1 (Gabry and Češnovar,
2023). We interpolated missing adaxial guard cell lengths from 6 out of
185 samples with zero adaxial stomata using the ``mi'' function in
\textbf{brms} package. We sampled the posterior distribution from 4
chains with 1000 iterations each after 1000 warmup iterations per chain.
From each posterior sample, we calculated \(g_{\mathrm{smax,ratio}}\)
as:

@@

Coastal ʻilima are surrounded by shorter vegetation than their montane
counterparts (\autoref{fig:study-system}d; Welch Two Sample
\emph{t}-test, \(t_{5.41} = 4.29\), \(P = 0.007\)). The montane site
with the lowest vegetation height is a remnant dry forest (Koaiʻa tree
sanctuary) in a matrix of cattle pasture, hence the satellite derived
vegetation height may be lower than what existed prior to human
disturbance. Coastal sites receive greater average solar radiation at
the top of the canopy (\autoref{fig:study-system}d; Welch Two Sample
\emph{t}-test, \(t_{9.93} = -1.89\), \(P = 0.089\)); coastal sites are
significantly warmer (\autoref{fig:study-system}d; Welch Two Sample
\emph{t}-test, \(t_{5.01} = -2.86\), \(P = 0.035\)); and coastal sites
receive less precipitation (\autoref{fig:study-system}d; Welch Two
Sample \emph{t}-test, \(t_{6.15} = 2.21\), \(P = 0.068\)).

@@

Amphistomy increases photosynthesis in leaves of coastal ʻilima plants
more than that of montane plants. \(\mathrm{AA}\) was significantly
greater than 0 (95\% confidence intervals did not overlap 0) in 5 of 6
coastal leaves, but only 1 of 6 montane leaves
(\autoref{fig:habitat-aa}; see \autoref{fig:licor} for individual
curves). Overall, the average \(\mathrm{AA}\) among coastal and montane
leaves is \(0.12~[\numrange{0.077}{0.15}]\) and
\(0.027~[\numrange{-0.0044}{0.057}]\), respectively; the difference in
average \(\mathrm{AA}\) between habitat types is
\(\mathrm{AA}_\text{coastal} - \mathrm{AA}_\text{montane} = 0.089~[\numrange{0.04}{0.14}]\).
Posterior predictions closely match observed values of \(A\)
(\autoref{fig:pp-licor}), indicating an adequate model fit from which we
can interpolate between measurements reliably. It also suggests that
slight extrapolation beyond the data should be reliable, but this is
less certain. When we remove two leaves where extrapolated slightly
beyond fitted \(A \textendash g_\mathrm{sw}\) curves, we estimate that
\(\mathrm{AA}_\text{coastal}\) is still positive,
\(0.081~[\numrange{0.022}{0.14}]\), but the difference between coastal
and montane leaves is smaller, \(0.053~[\numrange{-0.011}{0.12}]\), and
confidence intervals slightly overlap 0. Maximum photosynthetic rate was
slightly, but not significantly higher in coastal leaves (Welch Two
Sample \emph{t}-test, \(t_{9.65} = 1.6\), \(P = 0.14\)); total stomatal
conductance was similar (Welch Two Sample \emph{t}-test,
\(t_{9.71} = -0.09\), \(P = 0.93\)) in coastal and montane leaves
(\autoref{fig:habitat-Ags}). Water-use efficiency
(\(A / g_\mathrm{sw}\)) was significantly higher in coastal leaves
(Welch Two Sample \emph{t}-test, \(t_{9.99} = 2.54\), \(P = 0.03\)).

@@

Coastal ʻilima leaves are \(91~[\numrange{26}{164}]\) \(\mu\text{m}\)
thicker than their montane counterparts. Although coastal leaves are
thicker and have greater \(\mathrm{AA}\), there is little relationship
between leaf thickness and \(\mathrm{AA}\) within habitats
(\autoref{fig:traits-aa}A;
\(\text{slope}~= -0.11~[\numrange{-0.27}{0.036}]\)).

@@

Coastal and montane leaves have similar average
\(g_\mathrm{smax,ratio}\), the ratio of adaxial (upper) to total
\(g_\mathrm{smax}\), the anatomical maximum stomatal conductance to
water vapor (\autoref{fig:habitat-gmaxratio}); coastal leaves have
\(0.059~[\numrange{-0.14}{0.28}]\) higher \(g_\mathrm{smax,ratio}\) than
montane leaves, but the 95\% confidence intervals overlap 0 difference.
The \(g_\mathrm{smax,ratio}\) is somewhat bimodal among sites. Some
sites in both habitats have leaves with \(g_\mathrm{smax,ratio} < 0.07\)
and others with \(g_\mathrm{smax,ratio} > 0.2\)
(\autoref{fig:habitat-gmaxratio}). This is particularly noticeable in
montane sites where those on the Big Island of Hawaiʻi all have low
\(g_\mathrm{smax,ratio}\) whereas those on Oʻahu have relatively high
\(g_\mathrm{smax,ratio}\). There is no relationship between
\(g_\mathrm{smax,ratio}\) and \(\mathrm{AA}\) in either habitat
(\autoref{fig:traits-aa}B;
\(\text{slope}~= 0.14~[\numrange{-0.06}{0.35}]\)) in our sample.

@@

Amphistomy is a seemingly simple way that leaves can increase carbon
gain without significant additional water loss, yet it is rare in nature
and we do not know why. The strong association between amphiostomy and
sunny, open habitats suggests that amphistomy may benefit sun leaves
more than shade leaves, but progress has been limited by the lack of
evidence that amphiostomy \emph{per se} affects improves photosynthesis
in a given leaf. By experimentally blocking gas exchange through the
upper surface in a controlled environment, we directly compared an
amphistomatous leaf to an otherwise identical pseudohypostomatous leaf.
This allows us to quantity the amphistomy advantage (\(\mathrm{AA}\))
holding all else constant. Taking advantage of the steep climatic
gradients in the Hawaiian archipelago, we applied this new method to
show for the first time that sun leaves benefit 4 times more from
amphistomy than shade leaves on ʻilima (\emph{Sida fallax}) plants
(\(\mathrm{AA}_\text{coastal} = 0.12\)
vs.~\(\mathrm{AA}_\text{montane} = 0.027\)). Coastal and montane ʻilima
leaves are likely good representatives of classic sun and shade leaf
syndromes because 1) they vary in traits like reflective pubescence
(Ehleringer and Björkman, 1978) and leaf thickness (Terashima et al.,
2001) that typically characterize sun-shade adaptation; and 2) since
ʻilima shrubs are typically \(<1\)m tall, they are shaded by trees in
montane, but not coastal habitats (\autoref{fig:study-system}d). While
this result has not yet been validated in other species, our results
indicate that part of the reason amphistomatous leaves are found most
commonly in high light habitats is that the adaptive benefit is greater
in this environment.

If \(\mathrm{AA}\) is typically greater in sun leaves than shade leaves,
it could partially explain the distribution of amphi- and hypostoamtous
leaves, but the precise mechanism(s) require further study. One
hypothesis is that the internal airspace conductance,
\(g_\mathrm{ias}\), from stomata to mesophyll cell walls is lower in
thicker sun leaves (Parkhurst, 1978). All else being equal, a leaf with
lower \(g_\mathrm{ias}\) will benefit more from amphistomy. Our results
partially support this hypothesis. Coastal ʻilima leaves with high
\(\mathrm{AA}\) (\autoref{fig:habitat-aa}) are thicker than montane
leaves, but the relationship between \(\mathrm{AA}\) and leaf thickness
within habitats is actually slightly negative
(\autoref{fig:traits-aa}a), opposite our prediction. Since coastal and
montane leaves differ in many respects besides thickness, we do not have
enough data to conclude that leaf thickness explains the variation in
\(\mathrm{AA}\) between habitats. Alternatively, other biochemical or
anatomical differences between coastal and montane leaves may explain
why \(\mathrm{AA}\) is greater in coastal leaves. The negative
relationship, (albeit nonsignificant in that 95\% confidence intervals
encompassed 0), between leaf thickness and \(\mathrm{AA}\) could be
explained if thicker leaves compensated by having a more porous
mesophyll and/or less tortuous airspaces (Théroux-Rancourt et al.,
2021).

@@

Among land plants grown in a common garden, amphistomatous leaves have
on average nearly \(2\times\) higher area-based photosynthetic rates
(Xiong and Flexas, 2020), naively implying an
\(\mathrm{AA} \approx \text{log}~2 = 0.69\). This is much higher than
our estimate of 0.12 among coastal ʻilima leaves. The most likely
explanation is that amphistomy is not the only cause of high
photosynthetic rate. Indeed, species adapted to open, high light
habitats with amphistomatous leaves also have higher concentrations of
Rubisco, overall stomatal conductance, and photosynthetic capacity
(Smith et al., 1997; Xiong and Flexas, 2020). For a leaf with high
photosynthetic capacity that is well illuminated and hydrated, the major
limitation becomes CO\(_2\). Under these conditions, amphistomy may
substantially increase photosynthesis, as we observe in coastal ʻilima
leaves. Selection on increased photosynthesis under similar conditions
may explain why crops leaves tend to increase stomatal density ratio
during domestication (Milla et al., 2013).

Three limitations of this study are the small sample size, experimental
design that precludes distinguishing genetic from environmental
differences in leaf traits, and potentially confounding effects of other
environmental differences besides light environment. Understanding the
mechanistic basis of higher \(\mathrm{AA}\) in sun leaves would require
much larger sample sizes. Sun leaves tend to be thicker, more densely
packed with mesophyll cells, and have greater photosynthetic capacity
and higher stomatal conductance, among other traits (Lambers et al.,
2008). Each of these factors and others potentially modulate
\(\mathrm{AA}\). Quantifying the contribution of all these factors
requires larger samples and additional measurements that are beyond the
scope of this study, but exciting avenues for future research on leaf
structure-function relations. Although many morphological traits that
distinguish coastal and montane ʻilima populations persist in a common
environment (Yorkston and Daehler, 2006), we cannot distinguish between
genetic effects and plastic responses to habitat as causes of difference
in \(\mathrm{AA}\) because we measured naturally occurring plants
\emph{in situ}. While disentangling genetic and plastic contributions is
not necessarily important for understanding the distribution of
amphistomatous leaves, it would be insightful to know about genetic and
environmental contributions to trait variation. A reciprocal transplant
would be able to determine the genetic and environmental contributions,
as well their interaction, to trait variance in nature. However,
reciprocal transplants cannot control for other differences between
coastal and montane habitats besides vegetation height, such as
temperature and precipitaiton. Experimental studies in controlled
environments will be necessary to isolate the effect light quantity and
quality on \(\mathrm{AA}\).

>>>>

\begin{document}
\maketitle
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, enhanced, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, breakable, interior hidden, frame hidden]}{\end{tcolorbox}}\fi

@@

\textbf{Key results} Coastal ʻilima leaves benefit 4.04 times more from
amphistomy compared to their montane counterparts. Our evidence was
equivocal with respect to two hypotheses -- that coastal leaves benefit
more because 1) they are thicker and therefore have lower CO\(_2\)
conductance through the internal airspace, and 2) that they benefit more
because they have similar conductance on each surface, as opposed to
most of the conductance being on the lower (abaxial) surface.

\textbf{Conclusions} This is the first direct experimental evidence that
amphistomy \emph{per se} increases photosynthesis, consistent with the
hypothesis that parallel pathways through upper and lower mesophyll
increase the supply of CO\(_2\) to chloroplasts. The prevalence of
amphistomatous leaves in open, sunny habitats can partially be explained
the increased benefit of amphistomy in `sun' leaves, but the mechanistic
basis of this observation is an area for future research.

@@

\autoref{fig:ags-curve} illustrates our method for collecting
\(A \textendash g_\text{sw}\) curves. We collected two curves per leaf,
an amphi (untreated) curve and a pseudohypo (treated) curve. To control
for order effects, we alternated between starting with amphi or
pseudohypo leaf measurements, though we did not detect an effect of
treatment order on \(\mathrm{AA}\) (results not shown). In the field, we
acclimated the focal leaf to saturating light and high relative humidity
(\(\mathrm{RH} = 70\%\)), as described above, until \(A\) and
\(g_\text{sw}\) reach their maximum. We used these data as our estimates
of maximum \(A\) and \(g_\text{sw}\). After that, we decreased
\(\mathrm{RH}\) to \(\approx 10\%\) to induce rapid stomatal closure
without biochemical downregulation. Hence, \(A_\text{amphi}\) and
\(A_\text{hypo}\) were both measured at low chamber humidity after the
leaf had acclimated to high humidity. All other environmental conditions
in the leaf chamber remained the same. We logged data until
\(g_\text{sw}\) reached its nadir. We then repeated the process of
acclimating the leaf to 70\% \(\mathrm{RH}\) and inducing stomatal
closure with low \(\mathrm{RH}\) with the other treatment (amphi or
pseudohypo).

@@

In 10 of 12 leaves, the minimum \(g_\text{sw}\) of the amphi curve was
smaller than the maximum \(g_\text{sw}\) of the pseudohypo curve
(i.e.~the curves overlapped for a range of \(g_\text{sw}\) values). In
those cases, we estimated \(\hat{A}_\text{amphi}\) and
\(\hat{A}_\text{hypo}\) at the \(g_\mathrm{sw}\) value in the middle of
the range of overlap between the curves. In 2 of 12 leaves, the
\(A \textendash g_\text{sw}\) curves did not quite overlap because the
minimum \(g_\text{sw}\) of the amphi curve was slightly greater than the
maximum \(g_\text{sw}\) of the pseudohypo curve. In those cases, we
estimated \(\mathrm{AA}\) by extrapolating slightly,
\(1.98\times 10^{-3}~\text{and}~3.29\times 10^{-3}~\text{mol}~\text{m}^{-2}~\text{s}^{-1}\),
beyond the measured curves to the \(g_\mathrm{sw}\) value in between the
curves. The vertical lines in \autoref{fig:licor} show the
\(g_\text{sw}\) for each leaf. We estimated \(\mathrm{AA}\) from
\(\hat{A}_\text{amphi}\) and \(\hat{A}_\text{hypo}\) for each leaf using
the log-response ratio shown above.

To estimate \(\hat{\beta}\)'s from the \(A \textendash g_\text{sw}\)
curve for each leaf, we fit Bayesian regressions using the \emph{R}
package \textbf{brms} version 2.20.4 (Bürkner, 2017) with MCMC sampling
in \emph{Stan} (Stan Development Team, 2023). We used CmdStan version
2.33.1 and \textbf{cmdstanr} version 0.6.1 (Gabry and Češnovar, 2023) to
interface with \emph{R} version 4.3.1 (R Core Team, 2023). We sampled
the posterior distribution from 4 chains with 1000 iterations each after
1000 warmup iterations per chain. We estimated parameters and confidence
intervals as the median and 95\% quantile intervals of the posterior,
respectively. The key prediction is that
\(\mathrm{AA}_\text{coastal} > \mathrm{AA}_\text{montane}\), meaning the
95\% confidence intervals of
\(\mathrm{AA}_\text{coastal} - \mathrm{AA}_\text{montane}\) should be
positive and not encompass 0.

@@

We tested whether leaf thickness (log-transformed) varied between
coastal and montane populations and among individuals within populations
using a Bayesian mixed-effects model with habitat as a fixed effect and
individual plant and site as random effects. We used the \emph{R}
package \textbf{brms} version 2.20.4 (Bürkner, 2017) to fit the model in
\emph{Stan} (Stan Development Team, 2023) with CmdStan version 2.33.1
and \textbf{cmdstanr} version 0.6.1 (Gabry and Češnovar, 2023). We
sampled the posterior distribution from 4 chains with 1000 iterations
each after 1000 warmup iterations per chain. We estimated the
relationship between population average leaf thickness and
\(\mathrm{AA}\) measured from a single individual per population. We
used this approach because most of the variation in leaf thickness was
among sites and the plant selected for gas exchange measurements was not
always among the plants randomly selected for leaf thickness, precluding
individual level correlation. We propagated uncertainty about in
\(\mathrm{AA}\) and leaf thickness estimates by integrating over the
entire posterior distribution sample for each variable. The key
prediction is that the effect of leaf thickness on \(\mathrm{AA}\) is
positive, meaning the 95\% confidence interval of the slope should be
positive and not encompass 0.

@@

We tested whether \(g_\mathrm{smax,ratio}\) varied between coastal and
montane populations and among individuals within populations using a
Bayesian mutliresponse, mixed-effects model. The modeled response
variables are stomatal count and guard cell length on each surface.
Counts were modeled as negative binomially distributed variable from a
latent stomatal density and a parameter \(\phi\) to estimate
overdispersion in counts relative to a Poisson model. For all traits,
the explanatory variables were habitat as a fixed effect and leaf within
individual plant, individual plant, and site as random effects. We used
the \emph{R} package \textbf{brms} version 2.20.4 (Bürkner, 2017) to fit
the model in \emph{Stan} (Stan Development Team, 2023) with CmdStan
version 2.33.1 and \textbf{cmdstanr} version 0.6.1 (Gabry and Češnovar,
2023). We interpolated missing adaxial guard cell lengths from 6 out of
185 samples with zero adaxial stomata using the ``mi'' function in
\textbf{brms} package. We sampled the posterior distribution from 4
chains with 1000 iterations each after 1000 warmup iterations per chain.
From each posterior sample, we calculated \(g_{\mathrm{smax,ratio}}\)
as:

@@

Coastal ʻilima are surrounded by shorter vegetation than their montane
counterparts (\autoref{fig:study-system}d; Welch Two Sample
\emph{t}-test, \(t_{6.67} = 5.13\), \(P = 0.002\)). The montane site
with the lowest vegetation height is a remnant dry forest (Koaiʻa tree
sanctuary) in a matrix of cattle pasture, hence the satellite derived
vegetation height may be lower than what existed prior to human
disturbance. Coastal sites receive greater average solar radiation at
the top of the canopy (\autoref{fig:study-system}d; Welch Two Sample
\emph{t}-test, \(t_{10.86} = -2.22\), \(P = 0.049\)); coastal sites are
significantly warmer (\autoref{fig:study-system}d; Welch Two Sample
\emph{t}-test, \(t_{6.01} = -2.96\), \(P = 0.025\)); and coastal sites
receive less precipitation (\autoref{fig:study-system}d; Welch Two
Sample \emph{t}-test, \(t_{7.45} = 2.73\), \(P = 0.028\)).

@@

Amphistomy increases photosynthesis in leaves of coastal ʻilima plants
more than that of montane plants. \(\mathrm{AA}\) was significantly
greater than 0 (95\% confidence intervals did not overlap 0) in 5 of 6
coastal leaves, but only 1 of 6 montane leaves
(\autoref{fig:habitat-aa}; see \autoref{fig:licor} for individual
curves). Overall, the average \(\mathrm{AA}\) among coastal and montane
leaves is \(0.12~[\numrange{0.077}{0.15}]\) and
\(0.027~[\numrange{-0.0034}{0.057}]\), respectively; the difference in
average \(\mathrm{AA}\) between habitat types is
\(\mathrm{AA}_\text{coastal} - \mathrm{AA}_\text{montane} = 0.09~[\numrange{0.039}{0.14}]\).
Posterior predictions closely match observed values of \(A\)
(\autoref{fig:pp-licor}), indicating an adequate model fit from which we
can interpolate between measurements reliably. It also suggests that
slight extrapolation beyond the data should be reliable, but this is
less certain. When we remove two leaves where we extrapolated slightly
beyond fitted \(A \textendash g_\mathrm{sw}\) curves, we estimate that
\(\mathrm{AA}_\text{coastal}\) is still positive,
\(0.081~[\numrange{0.023}{0.13}]\), but the difference between coastal
and montane leaves is smaller, \(0.053~[\numrange{-0.012}{0.12}]\), and
confidence intervals slightly overlap 0. Maximum photosynthetic rate was
slightly, but not significantly higher in coastal leaves (Welch Two
Sample \emph{t}-test, \(t_{9.65} = 1.6\), \(P = 0.14\)); total stomatal
conductance was similar (Welch Two Sample \emph{t}-test,
\(t_{9.71} = -0.09\), \(P = 0.93\)) in coastal and montane leaves
(\autoref{fig:habitat-Ags}). Water-use efficiency
(\(A / g_\mathrm{sw}\)) was significantly higher in coastal leaves
(Welch Two Sample \emph{t}-test, \(t_{9.99} = 2.54\), \(P = 0.03\)).

@@

Coastal ʻilima leaves are \(91~[\numrange{26}{164}]\) \(\mu\text{m}\)
thicker than their montane counterparts. Although coastal leaves are
thicker and have greater \(\mathrm{AA}\), there is little relationship
between leaf thickness and \(\mathrm{AA}\) within habitats
(\autoref{fig:traits-aa}A;
\(\text{slope}~= -0.11~[\numrange{-0.28}{0.035}]\)).

@@

Coastal and montane leaves have similar average
\(g_\mathrm{smax,ratio}\), the ratio of adaxial (upper) to total
\(g_\mathrm{smax}\), the anatomical maximum stomatal conductance to
water vapor (\autoref{fig:habitat-gmaxratio}); coastal leaves have
\(0.059~[\numrange{-0.14}{0.28}]\) higher \(g_\mathrm{smax,ratio}\) than
montane leaves, but the 95\% confidence intervals overlap 0 difference.
The \(g_\mathrm{smax,ratio}\) is somewhat bimodal among sites. Some
sites in both habitats have leaves with \(g_\mathrm{smax,ratio} < 0.07\)
and others with \(g_\mathrm{smax,ratio} > 0.2\)
(\autoref{fig:habitat-gmaxratio}). This is particularly noticeable in
montane sites where those on the Big Island of Hawaiʻi all have low
\(g_\mathrm{smax,ratio}\) whereas those on Oʻahu have relatively high
\(g_\mathrm{smax,ratio}\). There is no relationship between
\(g_\mathrm{smax,ratio}\) and \(\mathrm{AA}\) in either habitat
(\autoref{fig:traits-aa}B;
\(\text{slope}~= 0.14~[\numrange{-0.057}{0.34}]\)) in our sample.

@@

Amphistomy is a seemingly simple way that leaves can increase carbon
gain without significant additional water loss, yet it is rare in nature
and we do not know why. The strong association between amphiostomy and
sunny, open habitats suggests that amphistomy may benefit sun leaves
more than shade leaves, but progress has been limited by the lack of
evidence that amphiostomy \emph{per se} improves photosynthesis in a
given leaf. By experimentally blocking gas exchange through the upper
surface in a controlled environment, we directly compared an
amphistomatous leaf to an otherwise identical pseudohypostomatous leaf.
This allows us to quantity the amphistomy advantage (\(\mathrm{AA}\))
holding all else constant. Taking advantage of the steep climatic
gradients in the Hawaiian archipelago, we applied this new method to
show for the first time that sun leaves benefit 4.04 times more from
amphistomy than shade leaves on ʻilima (\emph{Sida fallax}) plants
(\(\mathrm{AA}_\text{coastal} = 0.12\)
vs.~\(\mathrm{AA}_\text{montane} = 0.027\)). Coastal and montane ʻilima
leaves are likely good representatives of classic sun and shade leaf
syndromes because 1) they vary in traits like reflective pubescence
(Ehleringer and Björkman, 1978) and leaf thickness (Terashima et al.,
2001) that typically characterize sun-shade adaptation; and 2) since
ʻilima shrubs are typically \(<1\)m tall, they are shaded by trees in
montane, but not coastal habitats (\autoref{fig:study-system}d). While
this result has not yet been validated in other species, our results
indicate that part of the reason amphistomatous leaves are found most
commonly in high light habitats is that the adaptive benefit is greater
in such environments.

If \(\mathrm{AA}\) is typically greater in sun leaves than shade leaves,
it could partially explain the distribution of amphi- and hypostoamtous
leaves, but the precise mechanism(s) require further study. One
hypothesis is that the internal airspace conductance,
\(g_\mathrm{ias}\), from stomata to mesophyll cell walls is lower in
thicker sun leaves (Parkhurst, 1978). All else being equal, a leaf with
lower \(g_\mathrm{ias}\) will benefit more from amphistomy. Our results
partially support this hypothesis. Coastal ʻilima leaves with high
\(\mathrm{AA}\) (\autoref{fig:habitat-aa}) are thicker than montane
leaves, but the relationship between \(\mathrm{AA}\) and leaf thickness
within habitats is actually slightly negative
(\autoref{fig:traits-aa}a), opposite our prediction. Since coastal and
montane leaves differ in many respects besides thickness, we do not have
enough data to conclude that leaf thickness explains the variation in
\(\mathrm{AA}\) between habitats. Alternatively, other biochemical or
anatomical differences between coastal and montane leaves may explain
why \(\mathrm{AA}\) is greater in coastal leaves. The negative
relationship, albeit nonsignificant in that 95\% confidence intervals
encompassed 0, between leaf thickness and \(\mathrm{AA}\) could be
explained if thicker leaves compensated by having a more porous
mesophyll and/or less tortuous airspaces (Théroux-Rancourt et al.,
2021).

@@

Among land plants grown in a common garden, amphistomatous leaves have
on average nearly \(2\times\) higher area-based photosynthetic rates
(Xiong and Flexas, 2020), naively implying an
\(\mathrm{AA} \approx \text{log}~2 = 0.69\). This is much higher than
our estimate of 0.12 among coastal ʻilima leaves. The most likely
explanation is that amphistomy is not the only cause of high
photosynthetic rate. Indeed, species adapted to open, high light
habitats with amphistomatous leaves also have higher concentrations of
Rubisco, overall stomatal conductance, and photosynthetic capacity
(Smith et al., 1997; Xiong and Flexas, 2020). For a leaf with high
photosynthetic capacity that is well illuminated and hydrated, the major
limitation becomes CO\(_2\). Under these conditions, amphistomy may
substantially increase photosynthesis, as we observe in coastal ʻilima
leaves. Selection on increased photosynthesis under similar conditions
may explain why crop leaves tend to increase stomatal density ratio
during domestication (Milla et al., 2013).

Three limitations of this study are the small sample size, experimental
design that precludes distinguishing genetic from environmental
differences in leaf traits, and potentially confounding effects of other
environmental differences besides light environment. Understanding the
mechanistic basis of higher \(\mathrm{AA}\) in sun leaves would require
much larger sample sizes. Sun leaves tend to be thicker, more densely
packed with mesophyll cells, and have greater photosynthetic capacity
and higher stomatal conductance, among other traits (Lambers et al.,
2008). Each of these factors and others potentially modulate
\(\mathrm{AA}\). Quantifying the contribution of all these factors
requires larger samples and additional measurements that are beyond the
scope of this study, but exciting avenues for future research on leaf
structure-function relations. Although many morphological traits that
distinguish coastal and montane ʻilima populations persist in a common
environment (Yorkston and Daehler, 2006), we cannot distinguish between
genetic effects and plastic responses to habitat as causes of difference
in \(\mathrm{AA}\) because we measured naturally occurring plants
\emph{in situ}. While disentangling genetic and plastic contributions is
not necessarily important for understanding the distribution of
amphistomatous leaves, it would be insightful to know about genetic and
environmental contributions to trait variation. A reciprocal transplant
would be able to determine the genetic and environmental contributions,
as well their interaction, to trait variance in nature. However,
reciprocal transplants cannot control for other differences between
coastal and montane habitats besides vegetation height, such as
temperature and precipitaiton. Experimental studies in controlled
environments will be necessary to isolate the effects of light quantity
and quality on \(\mathrm{AA}\).

### 2023-11-03 ms/si.tex
revised ms

\begin{document}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, breakable, frame hidden, interior hidden, enhanced]}{\end{tcolorbox}}\fi

>>>>

\begin{document}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, sharp corners, interior hidden, enhanced, breakable]}{\end{tcolorbox}}\fi
